---
title: "Midterm SFO"
author: "Jim Turman"
date: "November 18, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
library(tufte)
library(leaps) # Exhaustive search for the best subsets of the variables in x for predicting y
library(e1071) # Skewness and Kurtosis
library(broom) # For residual analysis
library(ggplot2) # Plotting
library(sqldf) #to reformat the data
library(dplyr) #data prep
library(reshape2) #restructuring data
library(psych) #for descriptive stats
library(car) # levenes test
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

##Introduction
Below are the descriptive statistics, ANOVA and regression analysis of flight data for the SFO airport with regards to on time departure and arrival of flights. For the ANOVA testing a sample of 1000 observations is used where the departure time negatively effects the flight time meaning that there was a delay greater than zero. Attached with this file are plots describing the linear regression model that was done to predict the arrival time of flights leaving SFO.

##Descriptive Statistics 
###Descriptive Statistics for Airline Carriers
These are the descriptive statistics for the Airline Carriers and the delays associated with them. Here we can see that the airline carrier EV has the largest mean delay time (in minutes) as well as the largest median delay time which might suggest that they are the worst airline of the sample with regards to departure delay. This also shows that the Airline Carrier HA has the smallest average delay time of 15.058 minutes and a median of 6.5 minutes suggesting that they might have the the smallest average departure delay. Below those descriptive statistics are another set describing the Departure Delay as well as the types of delays that we have data for. The largest average delay time is due to late aircraft and the smallest average delay time is due to security delays. Following that block are the relative frequency of delays that have an effect on departure time (delay time is greater than zero) which would mean a late departure. From the given data set we can see that late aircraft delays represent the largest amount of delays with 2387 and security delays the least amount with 4.  
```{r airdesc, echo = FALSE}
setwd("C:/Users/jaturman/Desktop/804w/Midterm")
carrier.delay <- read.csv("DelayByCarrier")
setwd("C:/Users/jaturman/Desktop/data analysis/ops804/Midterm/data")
# Comment DK: You dont have to set the WD twice. Just set it once and use relative filepaths...

time <- read.csv("On_Time_On_Time_Performance_2016_8.csv")
delay <- sqldf("select DepDelay,DayOfWeek,UniqueCarrier,Origin,Dest,DestState,DepTime,Cancelled,CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, 
              LateAircraftDelay from time where DepDelayMinutes > 0 and Origin = 'SFO'")
# get data for most common delay 
delay.lr <- sqldf("select DepDelay, DayOfWeek, DepTime, Cancelled, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay
              FROM delay")
delay.carrier <- sqldf("select DepDelay,Carrier from time where DepDelayMinutes > 0 and Origin = 'SFO'")
# assign columns to vectors
aa <- c(carrier.delay$AA)
as <- c(carrier.delay$AS)
b6 <- c(carrier.delay$B6)
dl <- c(carrier.delay$DL)
oo <- c(carrier.delay$OO)
ua <- c(carrier.delay$UA)
vx <- c(carrier.delay$VX)
wn <- c(carrier.delay$WN)
ev <- c(carrier.delay$EV)
f9 <- c(carrier.delay$F9)
ha <- c(carrier.delay$HA)
nk <- c(carrier.delay$NK)
total <- c(aa,as,b6,dl,oo,ua,vx,wn,ev,f9,ha,nk)
n <- rep(1000,12)
group <- rep(c('AA','AS','B6','DL','OO','UA','VX','WN','EV','F9','HA','NK'),n)

# descriptive stats of data set for anova
tmp = function(x) c(sum=sum(x), mean = mean(x), median=median(x), mode = mode(x), var=var(x), sd = sd(x),
                    n = length(x), kurtosis=kurtosis(x),skew = skewness(x), max = max(x), min = min(x), range = max(x)-min(x), 
                    quartiles = quantile(x),iqr =IQR(x))

tapply(total,group,tmp)
describe(delay.lr)
# number of records for each delay type
    #number of carrier delays
count.cd = 0
for (i in delay$CarrierDelay){
  if (i != 0 && !is.na(i)){
    count.cd <- count.cd + 1
  }
}
cat("Number of carrier delays: ",count.cd)

    #number of weather delays
count.wd = 0
for (i in delay$WeatherDelay){
  if (i != 0 && !is.na(i)){
    count.wd <- count.wd + 1
  }
}
cat("Number of weather delays: ",count.wd)

  # number of NAS delays
count.nd = 0
for (i in delay$NASDelay){
  if (i != 0 && !is.na(i)){
    count.nd <- count.nd + 1
  }
}
cat("Number of NAS delays: ",count.nd)
  
  # number of security delays
count.sd = 0
for (i in delay$SecurityDelay){
  if (i != 0 && !is.na(i)){
    count.sd <- count.sd + 1
  }
}
cat("Number of security delays: ",count.sd)

# number of late aircraft delays
count.ad = 0
for (i in delay$LateAircraftDelay){
  if (i != 0 && !is.na(i)){
    count.ad <- count.ad + 1
  }
}
cat("Number of aircraft delays: ",count.ad)

# Comment DK: You also can calculate everything in an R file ouitside the markdown file making it more reuseable...
# Comment DK: You don't need the ANOVA, since we have population data here! 
```
##ANOVA
Below are the results of the ANOVA showing the difference between airline carriers with regards to the average negative delay time of their flights out of SFO. This analysis is testing to see if there is a statistically signifcant difference between the mean delay time of each carrier.  If there is a statistically significant difference between the mean delay time in minutes for each carrier we would expect to see a P value < .05. As we can see below the P value is much less than .05 so we can reject the null hypothesis that there is not a significant difference between the average delay times. There is also a plot showing how the data is distributed, because the data does not look like a normal bell curve we can assume that the sample of 1000 observations for each airline is not normally distributed.

```{r anova,echo=FALSE}
qqnorm(total)
#anova
data <- data.frame(total=total, group = factor(group))
fit <- lm(total~group, data)
carrier.delay.aov <- aov(fit)
carrier.delay.aov.summary <- summary(carrier.delay.aov)
print(carrier.delay.aov.summary)
```
###Levene's Test
Below is the Levene's test used to determine the homogeneity of variance between the airline carriers. My alpha level is .05 to determine the level of significance. If P is less than .05 I can conclude that I am 95% confident that the variance between the carriers is significantly different. The p value is far less than .05 so I can indeed conclude that there is a statistically significant difference between the variance of the carriers.
```{r levene,echo=FALSE}
#levenes test
carrier.lev <- leveneTest(total~group,data=data)
print(carrier.lev)
```
###Tukey-Kramer Post Hoc Test
This test is to determine which means have a difference between each other. The left hand column shows the two carriers that are being compared and the right hand shows the p value. the lower the P value the more significant the difference is between the average delay time of the carriers. My level of significance is .05. This suggests that the best performing airline in terms of average delay time is HA and the worst is EV.
```{r tukey, echo=FALSE}
# tukey-kramer
carrier.tukey <- TukeyHSD(carrier.delay.aov)
print(carrier.tukey)
plot(carrier.tukey)
```
##Regression Analysis
Below are the steps taken to develop a model to predict the arrival delay of a flight leaving SFO including flights that left both earlier and later than expected. The first step was to select fields that could possibly predict the arrival delay of a flight. I wanted to get as close to a 95%
[//]: # (Comment DK: I am not sure what you try to say here) 
confidence level as possible without making the model too flexible by adding too many fields. Below is an exhaustive search of the possible model for predicting the arrival delay. The higher the R-sq (adj) value the better the model will be at predicting arrival times. As we can see below the 10th iteration of the model produces the highest R-sq (adj) value and leaves us with greater than a 95% confidence level. 
[//]: # (Comment DK: R-sq(adj) stands for correlation - amount of explaind variability -, not for confidence intervall!)
That model includes the carrier, departure delay, taxi out time, distance traveled, air time, carrier delay, weather delay, NAS delay, security delay and late aircraft delay. 
```{r predmodel,echo=FALSE}
delay.log <- sqldf("select ArrDelay, case when Carrier = 'AA' then 1
                                    when Carrier = 'AS' then 2
                                    when Carrier = 'B6' then 3
                                    when Carrier = 'DL' then 4
                                    when Carrier = 'UA' then 5
                                    when Carrier = 'OO' then 6
                                    when Carrier = 'VX' then 7
                                    when Carrier = 'WN' then 8
                                    when Carrier = 'F9' then 9
                                    when Carrier = 'HA' then 10
                                    End as Carrier, DepDelay, TaxiOut, Distance, AirTime, 
                                    CarrierDelay, WeatherDelay,
                                    NASDelay, SecurityDelay,LateAircraftDelay
                    from time where Origin = 'SFO'")

# change all NA to 0 then write data to file for analysis
# comment out steps to build data frame after file created
delay.log[is.na(delay.log)] <- 0
DelayModel <- delay.log
rownames(DelayModel) <- NULL
write.csv(DelayModel, file = "DelayModelFields")


# read in formatted data for 
pred.mod <- read.csv("DelayModelFields")
model.dat <- pred.mod[-1]
x <- model.dat[,2:11]
y <- model.dat[,1]
model.out <- summary(regsubsets(x, y, nbest = 1, nvmax = ncol(x),force.in = NULL, force.out = NULL, method = "exhaustive"))
model.regtab <- cbind(model.out$which,model.out$rsq, model.out$adjr2, model.out$cp) # Stich things together
colnames(model.regtab) <- c("(Intercept)","Carrier","DepDelay","TaxiOut", "Distance", "AirTime", "CarrierDelay","WeatherDelay","NASDelay",
                            "SecurityDelay","LateAircraftDelay", "R-Sq", "R-Sq (adj)", "Cp") # Add header
print(model.regtab)
```
###Linear Regression
I have chosen to do a linear regression because the data is not categorical and is predicting a numerical value in terms of minutes. I am testing to see that at least one of the variables chosen contributes significantly to the model. If the P value is less than .05 then the variable is a significant contributor to the model. As shown below all of the variables contribute significantly and the overall P value of the model is 2.2e-16. The Adjusted R-squared is 0.9621 suggesting that with this model we can accurately predict arrival delay of flights leaving SFO with 96.21% confidence.
[//]: # (Comment DK: You did not test the assumptions! Are they all met?)
```{r linreg,echo=FALSE}
#linear regression
delay.lr <- lm(ArrDelay ~ ., data = model.dat)
delay.lr.summary <- summary(delay.lr) 
print(delay.lr.summary) # Show results
```

















